Contracts for unions and intersections are hard for Nickel

### 2021-06-01 paper.tex
Add an example of good error reporting

\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}

@@

Assertions are usually mainly used for correctness by enforcing pre- and
post-conditions of functions at runtime. In their foundational
paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
principled approach to run-time assertion checking called contracts. As opposed
to plain assertions, contracts can be composed naturally into more complex ones.
Nickel features such a contract system. For example, contracts nicely support
higher-order functions. If the user has written an \nickel{Even} contract for
numbers, they can use it right away to form a function contract \nickel{Even ->
Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
contract will check at each call that the argument is an even number
(pre-condition) and that the return value is a integer (post-condition). The
\nickel{#} character distinguishes contracts in Nickel. Contracts also introduce
the notion of blame which is crucial to good error reporting in presence of
higher-order contracts and other composed contracts.

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Even = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 2 == 0) in
let Int = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 1 == 0) in
{divBy2 | Even -> Int = fun x => x/2}
\end{lstlisting}
  \end{center}
\caption{Higher-order contract}
\label{fig:contract-higher-order}
\end{figure}

In our opinion, one of the main appeals of Nickel is this ability to write a
simple validation function for, say, a port number, and have it immediately
available to construct other contracts by writing simple type-like expressions.
Another example is given in Figure~\ref{fig:contract-composition}. The first
line defines a port contract from a predicate and use it in the definition of a
larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
contract checks that a value is a record with the following shape: it must have
a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
a \emph{schema} used to validate the final data, which is exactly what we were
looking for in the previous section. Not only it plays the role of a schema, but
it also \emph{looks} like one, as opposed to the corresponding plain validation
function. The \lstinline+|+ operator entails that a value respects a contract,
akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
been defined in a similar way. \unsure{What's missing in this section is an
    example of how contracts differ from standard dynamic type checking on some
function. Namely, the quality of error messages.}

\begin{figure}
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | Host,
  port | Port,
} in
{host = "nixos.org", port = 80}
| Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}

>>>>

\lstset{
  escapeinside={(*@}{@*)} % escape sequence to insert latex
}
\newcommand{\nickel}[1]{\lstinline[language=nickel]{#1}}
\newcommand{\racket}[1]{\lstinline[language=racket]{#1}}

@@

Assertions are usually mainly used for correctness by enforcing pre- and
post-conditions of functions at runtime. In their foundational
paper\cite{FindlerFelleisenHOContracts}, Findler and Felleisen introduce a
principled approach to run-time assertion checking called contracts. As opposed
to plain assertions, contracts can be composed naturally into more complex ones.
Nickel features such a contract system. For example, contracts nicely support
higher-order functions. If the user has written an \nickel{Even} contract for
numbers, they can use it right away to form a function contract \nickel{Even ->
Int} contract, as illustrated in Figure~\ref{fig:contract-higher-order}. This
contract will check at each call that the argument is an even number
(pre-condition) and that the return value is a integer (post-condition).
Contracts also introduce the notion of blame which is crucial to good error
reporting in presence of higher-order contracts and other composed contracts.

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=Nickel]
let Even = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 2 == 0) in
let Int = contracts.fromPred (fun val =>
  builtins.isNum val
  && val % 1 == 0) in
{divBy2 | Even -> Int = fun x => x/2}
\end{lstlisting}
  \end{center}
\caption{Higher-order contract}
\label{fig:contract-higher-order}
\end{figure}

In a dynamic language, even more so in a lazy one (see the paragraph on Lazyness
in Section~\ref{sec:feat-lang}), error messages may point to a location far
removed from the offending code. Contracts help by making assumptions on the
values passed at component boundaries explicit (components can be functions,
modules, classes, and so on, depending on language).

Figure~\ref{fig:contract-reporting-wo} and Figure~\ref{fig:contract-reporting-w}
give an example of a situation where adding a simple contract annotation
\nickel{Str -> Str} to a function greatly improves the error message . In the
original program of Figure~\ref{fig:contract-reporting-wo}, the
\nickel{catHosts} function takes a string argument, append it to a list that is
then folded. However, the programmer forgot one parameter when calling to
\nickel{makeHost}, thus passing a partially applied function instead of a string
to \nickel{catHosts}. Because this argument is put in a list and folded, the
error is raised later on and points to inside the code of \nickel{catHosts} and
\nickel{makeHost}, which may be totally unknown to the caller.

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel,title={Source program}]
let catHosts = fun last =>
  let hosts = ["foo.com", "bar.org"] in
  lists.fold (fun val acc => val ++ "," ++ acc) hosts last in

let makeHost = fun server ext => server ++ "." ++ ext in

catHosts (makeHost "google")
\end{lstlisting}

\begin{lstlisting}[title={Error message},frame=none,numbers=none]
error: Type error
3 | [(*@{\ldots}@*)] "," ++ acc) hosts last in
  |            ^^^
  | This expression has type Fun,
  | but Str was expected
4 |
5 | let mkHost = fun server ext => [(*@{\ldots}@*)] in
  |              -----------------------
  |              evaluated to this
  = ++, 2nd argument
\end{lstlisting}
  \end{center}
\caption{Error reporting without contract}
\label{fig:contract-reporting-wo}
\end{figure}

By adding a simple annotation to \nickel{catHosts} in
Figure~\ref{fig:contract-reporting-w}, the interpreter fails before even
entering the body of catHosts and is able to deliver a clearer message: the
first parameter of \nickel{catHosts} should be a string but the user provided
the body of the function \nickel{makeHost}. A stack trace is also shown,
including the faulty call to \nickel{catHost}.

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel,title={Added contract}]
let catHosts | Str -> Str = fun last =>
\end{lstlisting}

\begin{lstlisting}[title={Error message with contract},frame=none,numbers=none]
error: Blame error: contract broken by the caller.
  |  Str -> Str
  |  --- expected type of the argument [(*@{\ldots}@*)]
[(*@{\ldots}@*)]
1 | let catHosts | Str -> Str = fun last =>
  |                ^^^^^^^^^^ bound here
[(*@{\ldots}@*)]
6 | catHosts (makeHost "google")
  | ---------------------------- (2) calling <func>
\end{lstlisting}
  \end{center}
\caption{Error reporting with contract}
\label{fig:contract-reporting-w}
\end{figure}

The examples of Figure~\ref{fig:contract-reporting-wo} and
Figure~\ref{fig:contract-reporting-w} are about correctness. For validation, one
of the main appeals of Nickel in our opinion is the ability to write a simple
validation function for, say, a port number, and have it immediately available
to construct other contracts by writing simple type-like expressions.  Another
example is given in Figure~\ref{fig:contract-composition}. The first line
defines a port contract from a predicate and use it in the definition of a
larger \lstinline+Configuration+ record contract. The \lstinline+Configuration+
contract checks that a value is a record with the following shape: it must have
a \lstinline+port+ field obeying the \lstinline+Port+ contract, and a
\lstinline+host+ field obeying a \lstinline+Host+ contract. It plays the role of
a \emph{schema} used to validate the final data, which is exactly what we were
looking for in the previous section. Not only it plays the role of a schema, but
it also \emph{looks} like one, as opposed to the corresponding plain validation
function. The \lstinline+|+ operator entails that a value respects a contract,
akin to \lstinline+:+ for types. We assume that a \lstinline+Host+ contract has
been defined in a similar way. \resolved{What's missing in this section is an
example of how contracts differ from standard dynamic type checking on some
function. Namely, the quality of error messages.}

\begin{figure}[h]
  \begin{center}
\begin{lstlisting}[language=nickel]
let Port = contracts.fromPred (fun value =>
  builtins.isNum value && value % 1 == 0
  && value >= 0 && value <= 65353) in
let Configuration = {
  host | Host,
  port | Port,
} in
{host = "nixos.org", port = 80}
| Configuration
\end{lstlisting}
  \end{center}
\caption{Contract composition}
\label{fig:contract-composition}
\end{figure}
